Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-21 submission/manuscript.Rmd
add to methods

@@

>>>>

# Read in meta data
metaf <- read_csv("../data/raw/metadata/good_metaf_final.csv")
metai <- read_csv("../data/raw/metadata/metaI_final.csv")
geof_samples <- read_tsv("../data/raw/metadata/303030_subset/NexteraXT003Map.tsv", 
                         col_names = F) %>% filter(X22 != "Negative")

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before and after treatment for their tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol [@scfa_measures_venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations. These changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 in this report).

**16s rRNA gene sequencing.** The workflow and processing have been described previously [@Schloss2009; @Kozich2013; @normalization_sze2017]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity used the OptClust algorithm [@opticlust_westcott2017].

**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used. Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the normalize_by_copy_number.py function, and subsequent imputed metagenomes created using the predict_metagenomes.py function.

**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq with 125 bp paired end reads using a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF gene alignment used the KEGG database [@Kanehisa2015].

**Statistical analysis workflow.**

### 2018-05-21 submission/manuscript.tex
add to methods

@@

\hypertarget{refs}{}

>>>>

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (1, 2). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before and after treatment for their tumor (n=67). For patients
undergoing treatment for their tumor the length of time between their
initial and follow up sample ranged from 188 - 546 days. Our use of
treatment has been previously defined as encompassing removal of a tumor
with or without chemotherapy and radiation (1). Diagnosis of tumor was
made by colonoscopic examination and histopathological review of
biopsies obtained (1, 2). The University of Michigan Institutional
Review Board approved the study and informed consent was obtained from
all participants in accordance to the guidelines set out by the Helsinki
Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (3). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (3) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (1, 4, 5). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (6).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (7) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq with 125 bp paired end reads using a
previously described method (8). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (9). Open Reading Frames (ORFs) were
identified using Prodigal (10), counts generated using Diamond (11),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (12), and OPF gene alignment used the KEGG database (13).

\textbf{Statistical analysis workflow.}

@@

\hypertarget{refs}{}
\hypertarget{ref-normalization_sze2017}{}
1. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
2. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
3. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Schloss2009}{}
4. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
5. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
6. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Langille2013}{}
7. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

\hypertarget{ref-Hannigan2017}{}
8. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
9. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane K},
\textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution for
large and complex metagenomics assembly via succinct de bruijn graph.
Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
10. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
11. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
12. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
13. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.
